Tests of measurement invariance without subgroups: A~generalization of classical methods

### 2012-06-08 www/paper.Rnw
Final paper revisions: o Removed color from remaining figures, fixed text. o Added a citation based on some recent comments. o Spell check
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@136 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

%%      4. deal with color in figures
\title{Tests of measurement invariance without subgroups:
  A~generalization of classical methods}
\twoauthors{Edgar C. Merkle}{Achim Zeileis}
\twoaffiliations{University of Missouri}{Universit\"{a}t Innsbruck}

@@

\acknowledgements{This work was supported by National Science
  Foundation grant SES-1061334. The authors thank Jelte Wicherts, who
  generously shared data for the stereotype threat application,
  Yves Rosseel, who provided feedback and code for performing the
  tests with the lavaan package, and the participants of the
  Psychoco 2012 workshop on psychometric computing for helpful 
  discussion.
  Correspondence to Edgar C.\ Merkle, Department of  
  Psychological Sciences, University of Missouri, Columbia,
  MO 65211.
  Email: {\texttt{merklee@missouri.edu}}.}
\shorttitle{Tests of measurement invariance without subgroups}
\rightheader{Tests of measurement invariance without subgroups}

@@

However, taking maximums ``wastes'' power if many of the $k$~parameters change
at the same threshold, or if the score process takes large values for
many of the $n$~individuals (and not just a single threshold). In such
cases, sums insteads of maxima are more suitable for collapsing across
parameters and/or individuals, because they combine the deviations instead of
picking out only the single largest deviation. Thus, if the parameter instability
${\bm \theta}(V)$ affects many parameters and leads to many subgroups, sums
of (absolute or squared) values should be used for collapsing both across
parameters and individuals. On the other hand, if there is just a single
threshold that affects multiple parameters, then the natural aggregation
is by sums over parameters and then by the maximum over individuals.
More precisely, the former idea leads to a Cram{\'e}r-von Mises type
statistic and the latter to the maximum LM statistic from the previous section:
\begin{eqnarray}
    \label{eq:cvm}
    \mathit{CvM}     & = & n^{-1} \sum_{i = 1,\dots, n} \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2, \\
    \label{eq:maxlm}
    \max \mathit{LM} & = & \max_{i = \underline{i}, \dots, \overline{\imath}} ~
      \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}
      \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2,
\end{eqnarray}
where the $\max \mathit{LM}$ statistic is additionally scaled by the
asymptotic variance $t (1 - t)$ of the process ${\bm B}(t, \hat {\bm \theta})$.
It is equivalent to the $\max_\nu \mathit{LM}(\nu)$ statistic from the previous
section, provided that the boundaries for the subgroups sizes $\underline{i}$/$\underline{\nu}$
and $\overline{\imath}$/$\overline{\nu}$ are chosen analogously.
Further aggregation functions have been suggested in the structural
change literature (see e.g., \citeNP{Zei05,ZeiSha10}) but the three
tests above are most likely to be useful in psychometric settings.

Finally, all tests can be easily modified to address the situation of
so-called ``partial structural changes'' \cite{And93}. This refers to the case of some
parameters being known to be stable, i.e., restricted to be constant across
potential subgroups. Tests for potential changes/instabilities only in the
$k^*$ remaining parameters (from overall $k$ parameters) are easily obtained
by omitting those $k - k^*$ columns from ${\bm B}(\hat {\bm \theta})_{ij}$ that are
restricted/stable, retaining only those $k^*$ columns that are potentially instable.
This may be of special interest to those wishing to test specific types of measurement
invariance, where subsets of model parameters are assumed to be stable.

@@

As pointed out above, specification of the asymptotic distribution under $H_0$ for the test
statistics from the previous section is straightforward: it is simply the aggregation
of the asympotic process ${\bm B}^0(t)$ \cite{HjoKon02,ZeiHor07}. Thus, $\mathit{DM}$
from (\ref{eq:dmax}) converges to $\sup_t || {\bm B}^0(t) ||_{\infty}$, where
$||\cdot||_\infty$ denotes the maximum norm. Similarly, $\mathit{CvM}$
from (\ref{eq:cvm}) converges to $\int_0^1 || {\bm B}^0(t) ||_2^2 d t$, where $||\cdot||_2$
denotes the Euclidean norm. Finally, $\max \mathit{LM}$ from (\ref{eq:maxlm}) -- and analogously
the maximum Wald and LR~tests -- converges to $\sup_t (t (1-t))^{-1} || {\bm B}^0(t) ||_2^2$
(which can also be interpreted as the maximum of a tied-down Bessel process, as pointed out
previously).

@@

\subsection{Local Alternatives}
Using results from \citeA{HjoKon02} and \citeA{ZeiHor07}, we can also
capture the behaviour of the process $\bm{B}(t, \hat{\bm{\theta}})$
under specific alternatives 
of parameter instability. In particular, we can assume that the pattern
of deviation $\bm{\theta}(v_i)$ can be described as a constant
parameter plus some non-constant deviation $\bm{g}(i/n)$:
\begin{equation}
    \label{eq:local}
  \bm{\theta}(v_i) = \bm{\theta}_0 + n^{-1/2} \bm{g}(i/n).
\end{equation}
In this case, the scores $\bm{s}(\bm{\theta}_0; x_i)$ from
Equation~\eqref{eq:score} do not have zero expectation but rather
\begin{equation}
    \label{eq:localE}
  E[\bm{s}(\bm{\theta}_0; x_i)] = \bm{0} + n^{-1/2}
  \bm{C}(\bm{\theta}_0) \bm{g}(i/n).
\end{equation}
The covariance matrix $\bm{C}(\bm{\theta})$ from the expected
outer product of gradients is then 
\begin{equation}
    \label{eq:localS}
  \bm{C}(\bm{\theta}_0) = E[\bm{s}(\bm{\theta}_0; x_i) \bm{s}(\bm{\theta}; x_i)']    
\end{equation}
which, at $\bm{\theta}_0$, coincides with the expected information matrix.

@@

\subsection{Method}
WDH tested a series of confirmatory factor models for
various types of measurement invariance.  We focus on the model used
in their Step~5b (see pp.~703--704 of their paper), which involved
across-group restrictions on the factor loadings, unique
variances, intercepts, and factor variances.  These parameters were
typically restricted to be equal across groups, though a subset of the
parameters were allowed to be group-specific upon examination of
modification indices.    The model
provided a reasonable fit to the data, as judged by examination of
$\chi^2$, RMSEA (root-mean-square error of approximation), and CFI
(comparative fit index) statistics.  While the model included four
groups as described above, we focus only on the submodel for the minority
group that received stereotype primes, which is pictured in Figure~\ref{fig:wpath}.  In the figure, paths with dashed lines and \fixme{red} parameters signify group-specific parameters.  These include the factor mean $\eta$, numerical factor loading $\lambda_{\text{num}}$, numerical intercept $\mu_{\text{num}}$, and numerical unique variance $\psi_{\text{num}}$.  Other parameters (not displayed) were restricted to be equal across all four groups, with the exception of the factor variance.  The factor variance was restricted to be equal within both minority groups, and separately within both majority groups.

\begin{figure}
\caption{Path diagram representing the submodel for the ``minority, stereotype prime'' group in model 5b of WDH.  Dashed paths and red (FIXME) parameters are group specific.  Other parameters (not displayed) were restricted to be equal across all four groups, with the factor variance only being restricted across the two minority groups.}
\label{fig:wpath}
\includegraphics[height=3.5in]{wicherts_msp.pdf}
\end{figure}

@@

<<wdh_model>>=
## fit model 5b
wdh5b <- lavaan(
   'ability   =~ label(c(rep("load_n", 3), "load_n:min_t")) * numerical + label(rep("load_v", 4)) * verbal + 1 * abstract
    ability   ~  label(c(NA, "lmean:min_c", "lmean:maj_t", "lmean:min_t")) * 1 + c(0, NA, NA, NA) * 1
    ability   ~~ label(c("lvar:maj", "lvar:min", "lvar:maj", "lvar:min")) * ability
    numerical ~  label(c(rep("mean_n", 3), "mean_n:min_t")) * 1
    abstract  ~  label(c("mean_a:maj_c", "mean_a:min_c", rep("mean_a", 2))) * 1
    verbal    ~  label(rep("mean_v", 4)) * 1
    numerical ~~ label(c(rep("var_n", 3), "var_n:min_t")) * numerical
    abstract  ~~ label(rep("var_a", 4)) * abstract
    verbal    ~~ label(rep("var_v", 4)) * verbal',
  data = wdh, meanstructure = TRUE, group = "group") ##, likelihood = "wishart")

## refit model with new identification constraints
wdh5b_1 <- lavaan(
   'ability   =~ 1 * verbal + label(c(rep("load_n", 3), "load_n:min_t")) * numerical  + label(rep("load_a", 4)) * abstract
    ability   ~  label(c(NA, "lmean:min_c", "lmean:maj_t", "lmean:min_t")) * 1 + c(0, NA, NA, NA) * 1
    ability   ~~ label(c("lvar:maj", "lvar:min", "lvar:maj", "lvar:min")) * ability
    numerical ~  label(c(rep("mean_n", 3), "mean_n:min_t")) * 1
    abstract  ~  label(c("mean_a:maj_c", "mean_a:min_c", rep("mean_a", 2))) * 1
    verbal    ~  label(rep("mean_v", 4)) * 1
    numerical ~~ label(c(rep("var_n", 3), "var_n:min_t")) * numerical
    abstract  ~~ label(rep("var_a", 4)) * abstract
    verbal    ~~ label(rep("var_v", 4)) * verbal',
  data = wdh, meanstructure = TRUE, group = "group") #, likelihood = "wishart")
@

@@

\begin{figure}
\centering
\caption{Cumulative score processes for the four group-specific parameters across GPA
  in the ``minority, stereotype prime'' group
  using data from Study 1 of WDH. %% \citeA{WicDol05} does not work in caption
  The solid gray horizontal lines correspond to the critical value of
  the doule-max test at $\alpha = 0.05$.}
\label{fig:cusumgpa}
\setkeys{Gin}{width=0.8\textwidth}
<<gefp-4-plot, fig=TRUE, height=8, width=5.5>>=
gefp_4_bound <- get_boundary(gefp_4_info, maxBB)
plot(gefp_4_info$process,
  ylim = c(-2.15, 2.15), main = "DM, k* = 4", xlab = "GPA",
  ylab = expression(lambda[num], eta, mu[num], psi[num]),
  panel = function(x, y, ...) {
    abline(h = 0)
    lines(gefp_4_bound, col = "slategray", lwd = 1.7)
    lines(-gefp_4_bound, col = "slategray", lwd = 1.7)
    lines(x, y, ...)
  })

@@

The step functions from~\eqref{eq:h1*} were employed in the current paper to
highlight connections with nested model
comparison, but the proposed tests typically have non-trivial power for
all (non-constant) deviation patterns $\theta(v_i)$ (considered in
Equation~\eqref{eq:h1}). Thus, the tests will also have power for linear
deviations from parameter constancy, although other techniques may 
perform better in this particular case.  Examples include the 
class of moderated factor models \cite{BauHus09,MolDol10,Pur02}, 
whereby continuous moderators are allowed to linearly impact model
parameters (\citeNP{Pur02}, also considers quadratic effects of the
moderators). Moreover, if the change in
parameters is continuous but not exactly linear -- e.g., a sigmoidal
shift from one set of parameters to another one -- then the simple
single shift model in~\eqref{eq:h1*} may be a useful first
approximation and may have better power than linear techniques
(depending on how clearly the two regimes are separated).

>>>>

\title{Tests of measurement invariance without subgroups:
  A~generalization of classical methods}
\twoauthors{Edgar C. Merkle}{Achim Zeileis}
\twoaffiliations{University of Missouri}{Universit\"{a}t Innsbruck}

@@

\acknowledgements{This work was supported by National Science
  Foundation grant SES-1061334. The authors thank Jelte Wicherts, who
  generously shared data for the stereotype threat application,
  Yves Rosseel, who provided feedback and code for performing the
  tests with the lavaan package, Kris Preacher, who provided helpful
  comments 
  on the manuscript, and the participants of the
  Psychoco 2012 workshop on psychometric computing for helpful 
  discussion.
  Correspondence to Edgar C.\ Merkle, Department of  
  Psychological Sciences, University of Missouri, Columbia,
  MO 65211.
  Email: {\texttt{merklee@missouri.edu}}.}
\shorttitle{Tests of measurement invariance without subgroups}
\rightheader{Tests of measurement invariance without subgroups}

@@

However, taking maximums ``wastes'' power if many of the $k$~parameters change
at the same threshold, or if the score process takes large values for
many of the $n$~individuals (and not just a single threshold). In such
cases, sums instead of maxima are more suitable for collapsing across
parameters and/or individuals, because they combine the deviations instead of
picking out only the single largest deviation. Thus, if the parameter instability
${\bm \theta}(V)$ affects many parameters and leads to many subgroups, sums
of (absolute or squared) values should be used for collapsing both across
parameters and individuals. On the other hand, if there is just a single
threshold that affects multiple parameters, then the natural aggregation
is by sums over parameters and then by the maximum over individuals.
More precisely, the former idea leads to a Cram{\'e}r-von Mises type
statistic and the latter to the maximum LM statistic from the previous section:
\begin{eqnarray}
    \label{eq:cvm}
    \mathit{CvM}     & = & n^{-1} \sum_{i = 1,\dots, n} \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2, \\
    \label{eq:maxlm}
    \max \mathit{LM} & = & \max_{i = \underline{i}, \dots, \overline{\imath}} ~
      \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}
      \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2,
\end{eqnarray}
where the $\max \mathit{LM}$ statistic is additionally scaled by the
asymptotic variance $t (1 - t)$ of the process ${\bm B}(t, \hat {\bm \theta})$.
It is equivalent to the $\max_\nu \mathit{LM}(\nu)$ statistic from the previous
section, provided that the boundaries for the subgroups sizes $\underline{i}$/$\underline{\nu}$
and $\overline{\imath}$/$\overline{\nu}$ are chosen analogously.
Further aggregation functions have been suggested in the structural
change literature (see e.g., \citeNP{Zei05,ZeiSha10}) but the three
tests above are most likely to be useful in psychometric settings.

Finally, all tests can be easily modified to address the situation of
so-called ``partial structural changes'' \cite{And93}. This refers to the case of some
parameters being known to be stable, i.e., restricted to be constant across
potential subgroups. Tests for potential changes/instabilities only in the
$k^*$ remaining parameters (from overall $k$ parameters) are easily obtained
by omitting those $k - k^*$ columns from ${\bm B}(\hat {\bm \theta})_{ij}$ that are
restricted/stable, retaining only those $k^*$ columns that are potentially unstable.
This may be of special interest to those wishing to test specific types of measurement
invariance, where subsets of model parameters are assumed to be stable.

@@

As pointed out above, specification of the asymptotic distribution under $H_0$ for the test
statistics from the previous section is straightforward: it is simply the aggregation
of the asymptotic process ${\bm B}^0(t)$ \cite{HjoKon02,ZeiHor07}. Thus, $\mathit{DM}$
from (\ref{eq:dmax}) converges to $\sup_t || {\bm B}^0(t) ||_{\infty}$, where
$||\cdot||_\infty$ denotes the maximum norm. Similarly, $\mathit{CvM}$
from (\ref{eq:cvm}) converges to $\int_0^1 || {\bm B}^0(t) ||_2^2 d t$, where $||\cdot||_2$
denotes the Euclidean norm. Finally, $\max \mathit{LM}$ from (\ref{eq:maxlm}) -- and analogously
the maximum Wald and LR~tests -- converges to $\sup_t (t (1-t))^{-1} || {\bm B}^0(t) ||_2^2$
(which can also be interpreted as the maximum of a tied-down Bessel process, as pointed out
previously).

@@

\subsection{Local Alternatives}
Using results from \citeA{HjoKon02} and \citeA{ZeiHor07}, we can also
capture the behavior of the process $\bm{B}(t, \hat{\bm{\theta}})$
under specific alternatives 
of parameter instability. In particular, we can assume that the pattern
of deviation $\bm{\theta}(v_i)$ can be described as a constant
parameter plus some non-constant deviation $\bm{g}(i/n)$:
\begin{equation}
    \label{eq:local}
  \bm{\theta}(v_i) = \bm{\theta}_0 + n^{-1/2} \bm{g}(i/n).
\end{equation}
In this case, the scores $\bm{s}(\bm{\theta}_0; x_i)$ from
Equation~\eqref{eq:score} do not have zero expectation but rather
\begin{equation}
    \label{eq:localE}
  E[\bm{s}(\bm{\theta}_0; x_i)] = \bm{0} + n^{-1/2}
  \bm{C}(\bm{\theta}_0) \bm{g}(i/n).
\end{equation}
The covariance matrix $\bm{C}(\bm{\theta})$ from the expected
outer product of gradients is then 
\begin{equation}
    \label{eq:localS}
  \bm{C}(\bm{\theta}_0) = E[\bm{s}(\bm{\theta}_0; x_i) \bm{s}(\bm{\theta}; x_i)']    
\end{equation}
which, at $\bm{\theta}_0$, coincides with the expected information matrix.

@@

\subsection{Method}
WDH tested a series of confirmatory factor models for
various types of measurement invariance.  We focus on the model used
in their Step~5b (see pp.~703--704 of their paper), which involved
across-group restrictions on the factor loadings, unique
variances, intercepts, and factor variances.  These parameters were
typically restricted to be equal across groups, though a subset of the
parameters were allowed to be group-specific upon examination of
modification indices.    The model
provided a reasonable fit to the data, as judged by examination of
$\chi^2$, RMSEA (root-mean-square error of approximation), and CFI
(comparative fit index) statistics.  While the model included four
groups as described above, we focus only on the submodel for the minority
group that received stereotype primes, which is pictured in Figure~\ref{fig:wpath}.  In the figure, paths with dashed lines and parameters in bold/italics signify group-specific parameters.  These include the factor mean $\eta$, numerical factor loading $\lambda_{\text{num}}$, numerical intercept $\mu_{\text{num}}$, and numerical unique variance $\psi_{\text{num}}$.  Other parameters (not displayed) were restricted to be equal across all four groups, with the exception of the factor variance.  The factor variance was restricted to be equal within both minority groups, and separately within both majority groups.

\begin{figure}
\caption{Path diagram representing the submodel for the ``minority, stereotype prime'' group in model 5b of WDH.  Dashed paths and parameters in bold/italics are group specific.  Other parameters (not displayed) were restricted to be equal across all four groups, with the factor variance only being restricted across the two minority groups.}
\label{fig:wpath}
\includegraphics[height=3.5in]{wicherts_msp.pdf}
\end{figure}

@@

<<wdh_model>>=
## fit model 5b
wdh5b <- lavaan(
   'ability   =~ c(rep("load_n", 3), "load_n:min_t") * numerical + rep("load_v", 4) * verbal + 1 * abstract
    ability   ~  c(NA, "lmean:min_c", "lmean:maj_t", "lmean:min_t") * 1 + c(0, NA, NA, NA) * 1
    ability   ~~ c("lvar:maj", "lvar:min", "lvar:maj", "lvar:min") * ability
    numerical ~  c(rep("mean_n", 3), "mean_n:min_t") * 1
    abstract  ~  c("mean_a:maj_c", "mean_a:min_c", rep("mean_a", 2)) * 1
    verbal    ~  rep("mean_v", 4) * 1
    numerical ~~ c(rep("var_n", 3), "var_n:min_t") * numerical
    abstract  ~~ rep("var_a", 4) * abstract
    verbal    ~~ rep("var_v", 4) * verbal',
  data = wdh, meanstructure = TRUE, group = "group") ##, likelihood = "wishart")

## refit model with new identification constraints
wdh5b_1 <- lavaan(
   'ability   =~ 1 * verbal + c(rep("load_n", 3), "load_n:min_t") * numerical  + rep("load_a", 4) * abstract
    ability   ~  c(NA, "lmean:min_c", "lmean:maj_t", "lmean:min_t") * 1 + c(0, NA, NA, NA) * 1
    ability   ~~ c("lvar:maj", "lvar:min", "lvar:maj", "lvar:min") * ability
    numerical ~  c(rep("mean_n", 3), "mean_n:min_t") * 1
    abstract  ~  c("mean_a:maj_c", "mean_a:min_c", rep("mean_a", 2)) * 1
    verbal    ~  rep("mean_v", 4) * 1
    numerical ~~ c(rep("var_n", 3), "var_n:min_t") * numerical
    abstract  ~~ rep("var_a", 4) * abstract
    verbal    ~~ rep("var_v", 4) * verbal',
  data = wdh, meanstructure = TRUE, group = "group") #, likelihood = "wishart")
@

@@

\begin{figure}
\centering
\caption{Cumulative score processes for the four group-specific parameters across GPA
  in the ``minority, stereotype prime'' group
  using data from Study 1 of WDH. %% \citeA{WicDol05} does not work in caption
  The solid gray horizontal lines correspond to the critical value of
  the double-max test at $\alpha = 0.05$.}
\label{fig:cusumgpa}
\setkeys{Gin}{width=0.8\textwidth}
<<gefp-4-plot, fig=TRUE, height=8, width=5.5>>=
gefp_4_bound <- get_boundary(gefp_4_info, maxBB)
plot(gefp_4_info$process,
  ylim = c(-2.15, 2.15), main = "DM, k* = 4", xlab = "GPA",
  ylab = expression(lambda[num], eta, mu[num], psi[num]),
  panel = function(x, y, ...) {
    abline(h = 0)
    lines(gefp_4_bound, col = "slategray", lwd = 1.7)
    lines(-gefp_4_bound, col = "slategray", lwd = 1.7)
    lines(x, y, ...)
  })

@@

The step functions from~\eqref{eq:h1*} were employed in the current paper to
highlight connections with nested model
comparison, but the proposed tests typically have non-trivial power for
all (non-constant) deviation patterns $\theta(v_i)$ (considered in
Equation~\eqref{eq:h1}). Thus, the tests will also have power for linear
deviations from parameter constancy, although other techniques may 
perform better in this particular case.  Examples include the 
class of moderated factor models \cite{BauHus09,MolDol10,Nea06,Pur02}, 
whereby continuous moderators are allowed to linearly impact model
parameters (\citeNP{Pur02}, also considers quadratic effects of the
moderators). Moreover, if the change in
parameters is continuous but not exactly linear -- e.g., a sigmoidal
shift from one set of parameters to another one -- then the simple
single shift model in~\eqref{eq:h1*} may be a useful first
approximation and may have better power than linear techniques
(depending on how clearly the two regimes are separated).
